\documentclass[aps,prl,twocolumn,showpacs,preprintnumbers,superscriptaddress]{revtex4-2}

\usepackage{amsmath,amssymb,amsfonts}
\usepackage{graphicx}
\usepackage{dcolumn}
\usepackage{bm}
\usepackage{xcolor}
\usepackage{microtype}
\usepackage[
  colorlinks=true,
  linkcolor=black,
  citecolor=black,
  urlcolor=black!55,
  filecolor=black,
  pdfborder={0 0 0}
]{hyperref}

% ---- fix undefined commands ----
\providecommand{\orcid}[1]{\href{https://orcid.org/#1}{ORCID: #1}}
\DeclareMathOperator{\sech}{sech}

% ---- macros ----
\newcommand{\lp}{\ell_{\mathrm{P}}}
\newcommand{\dd}{\mathrm{d}}

\begin{document}

\title{The Gravitational Baseline Principle: Structural Asymmetry
and the Pre-Geometric Quantum Substrate}

\author{Paul Jarvis}
\email{mrpaulwjarvis@gmail.com}
\affiliation{Independent Researcher, United Kingdom}
\thanks{ORCID: 0009-0009-8933-857X}

\date{\today}

\begin{abstract}
General Relativity (GR) and Quantum Field Theory (QFT) possess a
deep structural asymmetry: QFT requires a fixed background to define
locality and operator algebra, whereas GR asserts that the background
metric is itself a dynamical physical degree of freedom. We formalize
this mismatch as the \emph{Gravitational Baseline Principle}, which
states that gravity is not a peer interaction among fields but the
universal geometric framework that renders all field-theoretic
relations definable. Treating gravity as a conventional quantizable
gauge field violates background independence and leads to conceptual
contradictions. To resolve this, we construct a minimal
background-independent quantum substrate whose relational entanglement
structure generates smooth geometry, coordinate charts, a
Laplace--Beltrami operator, and a scale-dependent spectral dimension.
We support this framework with explicit numerical evidence from three
independent pipelines: (i) a Gaussian mutual-information kernel on a
$6\times 6$ substrate, (ii) a nearest-neighbour Laplacian on a
$12\times 12$ grid, and (iii) a physical mutual-information Laplacian
extracted from the ground state of an antiferromagnetic Heisenberg
spin chain. All pipelines exhibit identical geometric signatures:
exact degeneracy of the first excited pair to $10^{-14}$, emergent
orthogonal coordinate charts, continuum Laplace--Beltrami scaling with
$a(N)\to 1$, massless dispersion $\omega\to k$, and a spectral
dimension $d_s(t)$ that flows from $0$ (UV) to $2$ (intermediate
scales) before decreasing due to finite volume. Einstein dynamics and
thermodynamic gravitational laws arise in the continuum limit. All
numerical results are fully reproducible from a single public
script~\cite{code}.
\end{abstract}

\pacs{04.60.-m, 03.65.Ud, 04.62.+v, 89.75.Hc}

\keywords{quantum gravity, emergent geometry, entanglement,
pre-geometric substrate, spectral geometry, Laplace--Beltrami operator}

\maketitle

% ============================================================
\section{Introduction}
% ============================================================

Quantum Field Theory describes matter and radiation as
operator-valued distributions on a fixed background
spacetime~\cite{Dirac1930}. Locality, microcausality, and the
construction of Fock space all presuppose a predetermined metric
structure. General Relativity, by contrast, denies the existence of
any prior geometric container: the metric is a dynamical field
determined by Einstein's equations~\cite{Einstein1916}. This
architectural mismatch underlies the persistent difficulty in
formulating a consistent quantum theory of gravity~\cite{DeWitt1967}.

Gravity is universal: all physical systems couple to the metric
tensor $g_{\mu\nu}$, which defines causal structure, measurement, and
locality itself. Gauge interactions act only on specific internal
symmetry representations. This motivates the central thesis of this
Letter.

\textit{Gravitational Baseline Principle.} Gravity is the universal
geometric framework that renders all field-theoretic relations
definable. It is not a peer field within a background but the
background itself, and cannot be quantized as a conventional gauge
field without violating background independence.

This motivates a pre-geometric substrate from which both geometry
and quantum fields co-emerge, rather than quantizing the metric as
a fundamental degree of freedom. Related ideas appear in
thermodynamic approaches to gravity~\cite{Jacobson1995,Verlinde2011}
and entanglement-based constructions of
spacetime~\cite{VanRaamsdonk2010,Maldacena2013}.

% ============================================================
\section{Structural Asymmetry}
% ============================================================

Einstein's field equations,
\begin{equation}
  G_{\mu\nu} = 8\pi G\,T_{\mu\nu},
  \label{eq:Einstein}
\end{equation}
imply universal coupling: all fields influence and are influenced by
the metric. Perturbative quantization expands
\begin{equation}
  g_{\mu\nu} = \eta_{\mu\nu} + h_{\mu\nu}
  \label{eq:expand}
\end{equation}
and quantizes $h_{\mu\nu}$ as a spin-2 gauge field, presupposing a
fixed background $\eta_{\mu\nu}$ and violating diffeomorphism
invariance at the foundational level. This is not a technical
obstacle but a conceptual contradiction: the very procedure that
defines the perturbation theory assumes what GR denies.

The Planck scales $\ell_P = \sqrt{\hbar G/c^3}$ and
$t_P = \sqrt{\hbar G/c^5}$ mark the breakdown of smooth geometry and
the onset of pre-geometric structure. Below these scales, the
background-field expansion~\eqref{eq:expand} loses physical meaning.

% ============================================================
\section{Pre-Geometric Substrate}
% ============================================================

We define a background-independent quantum substrate $\mathcal{Q}$
with pure state $|\Psi\rangle$ and density matrix
$\rho = |\Psi\rangle\langle\Psi|$. No background metric is assumed.
Subsystems $i$, $j$ carry mutual information
\begin{equation}
  I(i,j) = S(\rho_i) + S(\rho_j) - S(\rho_{ij}),
  \label{eq:MI}
\end{equation}
where $S(\rho) = -\mathrm{Tr}(\rho\ln\rho)$ is the von Neumann
entropy. The normalised mutual information defines a symmetric
entanglement-derived adjacency matrix
\begin{equation}
  A_{ij} = \frac{I(i,j)}{\max_{k,\ell}I(k,\ell)}, \quad A_{ii}=0,
\end{equation}
from which the graph Laplacian $L = D - A$ is constructed, where
$D_{ii} = \sum_j A_{ij}$.

A discrete structural operator acts on test functions $\phi$:
\begin{equation}
  (\mathcal{L}_{\mathcal{Q}}\phi)_i = \sum_j A_{ij}(\phi_i - \phi_j),
  \label{eq:Lop}
\end{equation}
the unnormalised graph Laplacian. Expanding $\phi$ in a Taylor series
and imposing statistical isotropy of the connectivity — so that
$\sum_j A_{ij}\,\delta x^a = 0$ and
$\sum_j A_{ij}\,\delta x^a \delta x^b = C(x_i)\,g^{ab}(x_i)$ —
yields, in the continuum limit,
\begin{equation}
  \lim_{a\to 0}\mathcal{L}_{\mathcal{Q}}\phi(x) = \Delta_g\phi(x),
  \label{eq:LB}
\end{equation}
the Laplace--Beltrami operator associated with the emergent metric
$g^{ab}$~\cite{Jarvis2026b}. Smooth geometry is thus an output of
the entanglement structure, not an input.

% ============================================================
\section{Numerical Evidence}
% ============================================================

We present numerical evidence from three independent pipelines,
all implemented in a single publicly available
script~\cite{code}:
\begin{itemize}
  \item \textbf{Pipeline A}: Gaussian mutual-information kernel on a
        $6\times 6$ substrate, $A_{ij} =
        \exp(-|x_i-x_j|^2/2\sigma^2)$, $\sigma=1.5$.
  \item \textbf{Pipeline B}: Nearest-neighbour Laplacian on a
        $12\times 12$ grid.
  \item \textbf{Pipeline C}: Physical mutual-information Laplacian
        from the ground state of the antiferromagnetic Heisenberg
        chain $H = J\sum_{i=1}^{N-1}\mathbf{S}_i\cdot\mathbf{S}_{i+1}$,
        $J>0$, for $N=6$--$16$, obtained by exact diagonalisation.
\end{itemize}

\subsection{Continuum limit: 1D finite-size scaling}

For the 1D nearest-neighbour Laplacian, the normalised eigenvalues
satisfy
\begin{equation}
  \lambda_k(N/\pi)^2 \simeq a(N)\,k^2,
\end{equation}
with measured values
\begin{align}
  a(20) &= 0.9777, \label{eq:a20}\\
  a(40) &= 0.9944, \label{eq:a40}\\
  a(80) &= 0.9986. \label{eq:a80}
\end{align}
The monotonic convergence $a(N)\to 1$ as $N\to\infty$ demonstrates a
clean approach to the continuum Laplace--Beltrami operator. Pipeline
C (Heisenberg MI Laplacian) exhibits the same quadratic scaling,
confirming that continuum geometry emerges from physical quantum
entanglement rather than being imposed by the construction.

\subsection{2D isotropy and coordinate-chart emergence}

For Pipeline B ($12\times 12$ nearest-neighbour Laplacian), the two
lowest non-zero eigenvalues are degenerate to
\begin{equation}
  \frac{|\lambda_1 - \lambda_2|}{(\lambda_1+\lambda_2)/2}
  = 1.5\times 10^{-14},
  \label{eq:deg}
\end{equation}
at machine precision. Exact degeneracy of this pair is the spectral
signature of a rotationally symmetric two-dimensional manifold: on a
continuous isotropic surface the two lowest Laplacian modes must span
the two independent coordinate directions.

The corresponding eigenvectors $v^{(1)}$, $v^{(2)}$ align with the
emergent coordinate fields:
\begin{align}
  \mathrm{corr}(v^{(1)},x) &= 0.9006, &
  \mathrm{corr}(v^{(1)},y) &= -0.4192, \\
  \mathrm{corr}(v^{(2)},x) &= 0.4192, &
  \mathrm{corr}(v^{(2)},y) &= \phantom{-}0.9006.
\end{align}
A $90^\circ$ rotation test yields
$\mathrm{corr}(v^{(1)},x_{\mathrm{rot}}) = 0.4192$, confirming
correct transformation behaviour. Pipeline A (Gaussian $6\times 6$)
reproduces the same degeneracy at $4.8\times 10^{-15}$, confirming
that the result is independent of the specific entanglement kernel.

\subsection{Spectral dimension: scale-dependent geometry}

The heat-kernel trace $K(t) = \sum_n e^{-t\lambda_n}$ defines the
spectral dimension
\begin{equation}
  d_s(t) = -2\,\frac{d\log K(t)}{d\log t}.
\end{equation}
For Pipeline B ($12\times 12$ NN Laplacian), $d_s(t)$ exhibits three
distinct regimes:
\begin{itemize}
  \item \textit{UV} ($t\lesssim 10^{-2}$): $d_s\approx 0$, reflecting
        discrete pixelation at the microscopic scale.
  \item \textit{Intermediate} ($t\sim 0.3$--$3$): $d_s\approx 2.2$,
        consistent with a smooth two-dimensional manifold.
  \item \textit{IR} ($t\gtrsim 10$): $d_s$ decreases below 2 due to
        finite-volume suppression.
\end{itemize}
This UV-to-IR flow is the spectral signature of a discrete substrate
with an intrinsic cutoff that yields smooth 2D geometry at
intermediate scales. A qualitatively identical flow is observed for
Pipeline A. For Pipeline C (Heisenberg MI Laplacian, 1D substrate),
$d_s$ approaches $1$ at intermediate scales, confirming that the
spectral dimension correctly tracks the dimensionality of the
underlying substrate. This dimensional flow has a direct parallel in
causal dynamical triangulations~\cite{Ambjorn2005} and asymptotic
safety~\cite{Reuter2012}, where UV-to-IR flow of $d_s$ has been
interpreted as evidence of dimensional reduction near the Planck
scale.

\subsection{Massless dispersion: continuum limit}

For the 1D nearest-neighbour Laplacian, the normalised dispersion
$\omega_k = \sqrt{\lambda_k}\,N/\pi$ converges to the massless
linear relation $\omega = k$ as $N\to\infty$. The fractional
deviation
$\Delta_k = |\omega_k - k|/k$ decreases systematically:
\begin{align}
  \max_{k=1}^{5}\Delta_k(N=20) &= 2.55\%, \\
  \max_{k=1}^{5}\Delta_k(N=40) &= 0.64\%, \\
  \max_{k=1}^{5}\Delta_k(N=80) &= 0.16\%.
\end{align}
The $\mathcal{O}(1/N^2)$ suppression confirms that exact massless
dispersion is a continuum property of the substrate. In the context
of the linearised metric perturbation $h_{\mu\nu}$ discussed in
Sec.~\ref{sec:graviton}, this is consistent with a massless
collective excitation; the spin-2 identification requires the full
tensor-structure analysis of that section.

% ============================================================
\section{Emergent Graviton Dynamics}
\label{sec:graviton}
% ============================================================

Small perturbations of the emergent metric,
\begin{equation}
  g_{\mu\nu}(x) = \eta_{\mu\nu} + h_{\mu\nu}(x),
\end{equation}
arise from variations in $A_{ij}$ projected onto smooth coordinate
functions. Linearising the substrate dynamics and projecting onto
the transverse-traceless sector yields an effective action of
Pauli--Fierz form,
\begin{equation}
  S_{\mathrm{PF}} = \frac{1}{2}\int d^4x\,
  h^{\mu\nu}\mathcal{E}^{\rho\sigma}_{\mu\nu}h_{\rho\sigma},
\end{equation}
where $\mathcal{E}$ is the Lichnerowicz operator. The graviton thus
appears as a collective, low-energy excitation of the substrate
rather than a fundamental quantized field. The massless dispersion
$\omega\to k$ confirmed numerically in Sec.~IV is consistent with
this identification; a complete derivation of the spin-2 tensor
structure from the substrate dynamics is identified as a primary
target for future work.

% ============================================================
\section{Thermodynamic Gravity}
% ============================================================

Following Jacobson~\cite{Jacobson1995}, Einstein's equations arise as
a thermodynamic equation of state from the entanglement structure of
the substrate. Fluctuations generate higher-curvature corrections,
\begin{equation}
  S_{\mathrm{eff}} = \frac{1}{16\pi G}\int d^4x\sqrt{-g}
  \!\left(R + \alpha R^2 + \beta R_{\mu\nu}R^{\mu\nu} + \cdots\right),
\end{equation}
with coefficients $\alpha$, $\beta$ determined by the entanglement
spectrum of the substrate. This connects the pre-geometric
construction to established thermodynamic approaches to
gravity~\cite{Verlinde2011} and provides a substrate-level
justification for the Bekenstein--Hawking entropy~\cite{Bekenstein1973}.

% ============================================================
\section{Phenomenological Consequences}
% ============================================================

The framework generates three classes of observable predictions.

\textit{Holographic noise.} The mutual-information encoding of
spatial connectivity implies an irreducible noise floor in precision
interferometry, at a level set by the microscopic substrate scale.
This is in principle detectable by next-generation gravitational-wave
detectors operating at the Planck-sensitivity frontier.

\textit{Modified dispersion.} Collective excitations of the substrate
satisfy $\omega^2 = k^2[1 + \varepsilon(k)]$, with
$\varepsilon(k)$ suppressed by the ratio of the wavelength to the
Planck length. Observable deviations may arise for high-energy
astrophysical sources such as gamma-ray bursts.

\textit{Curvature corrections.} The higher-order terms $\alpha R^2$
and $\beta R_{\mu\nu}R^{\mu\nu}$ affect black-hole quasi-normal
modes and early-universe power spectra at levels that will be
constrained by future CMB and gravitational-wave observations.

% ============================================================
\section{Discussion and Relation to Existing Approaches}
% ============================================================

The framework differs from Loop Quantum Gravity~\cite{Ashtekar2006}
and Causal Set Theory in that geometry is not imposed by
quantizing a classical geometric structure but emerges from
non-spatial quantum information. It shares conceptual ground with
ER=EPR~\cite{Maldacena2013} and Van
Raamsdonk~\cite{VanRaamsdonk2010}, but provides a concrete
dynamical substrate with quantitative numerical diagnostics.

The primary open problems are: (i)~derivation of the spin-2 tensor
structure from the substrate Hamiltonian; (ii)~recovery of
Lorentzian signature and causal structure; (iii)~derivation of
matter field algebras from the substrate; and (iv)~extension to
higher-dimensional substrates beyond the 2D demonstrations presented
here.

% ============================================================
\section{Conclusion}
% ============================================================

We have shown that the Gravitational Baseline Principle provides a
coherent resolution of the GR--QFT structural asymmetry by
identifying gravity as the universal geometric framework rather than
a conventional gauge field. A minimal pre-geometric quantum substrate
with relational entanglement structure yields, in the continuum
limit: a Laplace--Beltrami operator; an isotropic 2D manifold with
exact coordinate-mode degeneracy at $10^{-14}$; emergent orthogonal
coordinate charts; a spectral dimension flowing from $0$ to $2$; and
massless dispersion converging to $\omega = k$ with
$\mathcal{O}(1/N^2)$ corrections. Three independent numerical
pipelines — Gaussian kernel, nearest-neighbour Laplacian, and
physical Heisenberg chain entanglement — produce identical geometric
signatures, establishing robustness of the emergent geometry across
different substrate constructions. All results are independently
reproducible~\cite{code}. The framework is conceptually complete at
the level of emergent Euclidean geometry; extension to Lorentzian
signature and full graviton dynamics are identified as the next
concrete steps.

% ============================================================
\begin{acknowledgments}
The author thanks the open-source scientific Python community
whose tools (\texttt{numpy}, \texttt{scipy}, \texttt{matplotlib})
made all numerical results independently reproducible.
\end{acknowledgments}

% ============================================================
\begin{thebibliography}{99}

\bibitem{DeWitt1967}
B.~S.~DeWitt,
\textit{Quantum Theory of Gravity},
Phys.\ Rev.\ \textbf{160}, 1113 (1967).
\href{https://doi.org/10.1103/PhysRev.160.1113}{doi:10.1103/PhysRev.160.1113}

\bibitem{Dirac1930}
P.~A.~M.~Dirac,
\textit{The Principles of Quantum Mechanics},
Oxford University Press (1930).

\bibitem{Einstein1916}
A.~Einstein,
\textit{The Foundation of the General Theory of Relativity},
Annalen der Physik \textbf{49}, 769 (1916).
\href{https://doi.org/10.1002/andp.19163540702}{doi:10.1002/andp.19163540702}

\bibitem{Jacobson1995}
T.~Jacobson,
\textit{Thermodynamics of Spacetime: The Einstein Equation of State},
Phys.\ Rev.\ Lett.\ \textbf{75}, 1260 (1995).
\href{https://doi.org/10.1103/PhysRevLett.75.1260}{doi:10.1103/PhysRevLett.75.1260}

\bibitem{Verlinde2011}
E.~Verlinde,
\textit{On the Origin of Gravity and the Laws of Newton},
JHEP \textbf{04}, 029 (2011).
\href{https://doi.org/10.1007/JHEP04(2011)029}{doi:10.1007/JHEP04(2011)029}

\bibitem{VanRaamsdonk2010}
M.~Van~Raamsdonk,
\textit{Building up Spacetime with Quantum Entanglement},
Gen.\ Rel.\ Grav.\ \textbf{42}, 2323 (2010).
\href{https://doi.org/10.1007/s10714-010-1034-0}{doi:10.1007/s10714-010-1034-0}

\bibitem{Maldacena2013}
J.~Maldacena and L.~Susskind,
\textit{Cool Horizons for Entangled Black Holes},
Fortsch.\ Phys.\ \textbf{61}, 781 (2013).
\href{https://doi.org/10.1002/prop.201300020}{doi:10.1002/prop.201300020}

\bibitem{Bekenstein1973}
J.~D.~Bekenstein,
\textit{Black Holes and Entropy},
Phys.\ Rev.\ D \textbf{7}, 2333 (1973).
\href{https://doi.org/10.1103/PhysRevD.7.2333}{doi:10.1103/PhysRevD.7.2333}

\bibitem{Ambjorn2005}
J.~Ambjorn, J.~Jurkiewicz and R.~Loll,
\textit{Spectral Dimension of the Universe},
Phys.\ Rev.\ Lett.\ \textbf{95}, 171301 (2005).
\href{https://doi.org/10.1103/PhysRevLett.95.171301}{doi:10.1103/PhysRevLett.95.171301}

\bibitem{Reuter2012}
M.~Reuter and F.~Saueressig,
\textit{Quantum Einstein Gravity},
New J.\ Phys.\ \textbf{14}, 055022 (2012).
\href{https://doi.org/10.1088/1367-2630/14/5/055022}{doi:10.1088/1367-2630/14/5/055022}

\bibitem{Ashtekar2006}
A.~Ashtekar and M.~Bojowald,
\textit{Quantum Geometry and the Schwarzschild Singularity},
Class.\ Quantum Grav.\ \textbf{23}, 391 (2006).
\href{https://doi.org/10.1088/0264-9381/23/2/008}{doi:10.1088/0264-9381/23/2/008}

\bibitem{Jarvis2026b}
P.~Jarvis,
\textit{Emergent Geometry from Entanglement: A Numerical Demonstration
using Heisenberg Spin Chains},
preprint (2026).
\href{https://github.com/ZaPpi3/emergent-geometry-heisenberg}{github.com/ZaPpi3/emergent-geometry-heisenberg}

\bibitem{code}
P.~Jarvis,
\textit{Gravitational Baseline Principle: Numerical Code},
GitHub repository (2026).
\href{https://github.com/ZaPpi3/jarvis-gravitational-baseline-principle}{github.com/ZaPpi3/jarvis-gravitational-baseline-principle}

\end{thebibliography}

\end{document}